---
title: "PCA"
---

## Principal Component Analysis

PCA is a great way to explore your sequencing data, assess structure, patterns, or discover plate inversions. Yay. Generally it reduces the dimensionality of data and provides a way to view that data along two axes.

### IBS PCA

The newest `angsd` version (**0.918+**) provides a way to do PCA without subsampling your data. This creates a covariance matrix bt sampling a single read at each polymorphic site. The basic angsd call for this is as follows:

```{bash, eval=F, echo=T}
~/bin/angsd -bam ${bamlist} -out ${out}.PCAMDS1 -doIBS 1 -doCounts 1 -doMajorMinor 1 -minFreq 0.05 -maxMis ${nind} -minMapQ 30 -minQ 20 -SNP_pval 0.000001 -makeMatrix 1 -doCov 1 -GL 1 -doMaf 2 -doPost 2 -nThreads 16

```

**Notes**: 

 - when running with `doIBS` give extra time/memory. This method can take a long time because it uses all sites in the data.
 - if you've already run the subsample script, one option is to `ls` all bamfiles that have a certain subsampling threshold (say 100k), and then `sed` out the 100k part of the filename, so you match the the original bam file name, but only keep bams that have read counts 100k or greater. This is a good way to use all the data, but only for samples that have great than a certain number of reads.
 - The post-processing scripts for plotting should be largely the same.

### `ngsCovar` PCA with subsampling

This method uses the other genotype calling option and `ngsCovar`.

 > **PCA_CALC.sh**:

```{bash pca_calc, eval=F}

#!/bin/bash -l

#SBATCH -o slurm_outs/03_pca_calc-%j.out
#SBATCH -J pca_calc_sites

echo "Starting Job: "
date

bamlist=$1
out=$2
bamdir=$3 

nInd=$(wc ${bamdir}/${bamlist} | awk '{print $1}')
minInd=$[$nInd/2]
outdir='angsd_output'

angsd -bam ${bamdir}/${bamlist} -out ${outdir}/${out} -minQ 20 -minMapQ 10 -minInd $minInd -GL 1 -doMajorMinor 1 -doMaf 2 -doPost 2 -doGeno 32 -minMaf 0.05 -postCutoff 0.95 -SNP_pval 1e-6 -sites bait_lengths.txt

gunzip ${outdir}/${out}.*.gz

count=$(sed 1d ${outdir}/${out}.mafs | wc -l | awk '{print $1}')

ngsCovar -probfile ${outdir}/${out}.geno -outfile ${outdir}/${out}.covar -nind $nInd -nsites $count -call 1

echo "Ending Job: "
date

```


To plot, you'll need to assign IDs using a `clst` file. I use the following steps (assuming we are using the `100k` subsampled bamlist):

 1. Use `sed` to create a clst file of the samples/bamnames (i.e., `sed 's/_RA\.sort\.flt_100000\.bam/ 1 1/g' subbamlist > out.clst`)
 2. VIM (or other text editor) to edit the `out.clst` and insert a header **`FID IID CLUSTER`**:
    - (IID changes symbol; CLUSTER changes color)
    - Feel free to modify this file according to symbol or color as you see fit for the final PCA plot
 3. Use `pca_plot.sh` to create 3 pdfs (pc1 vs pc2, pc1 v pc3, pc2 vs. pc3). It can be modified.

**PCA_PLOT**
```{bash pca_plot_sh, eval=F}

#!/bin/bash -l

#SBATCH -o slurm_outs/04_pca_pdf-%j.out
#OUTDIR=pca_pdfs
#SBATCH -J PCA_plots

echo "Starting Job: "
date

out=$1
bamclst=$2

outdir='pca_pdfs'
homedir='/home/fastq'

Rscript --vanilla --slave  ${homedir}/pca_makeplot.R -i ./angsd_output/${out}.covar -c 1-2 -a ${homedir}/${bamclst} -o ${outdir}/${out}_12_pca.pdf
Rscript --vanilla --slave ${homedir}/pca_makeplot.R -i ./angsd_output/${out}.covar -c 1-3 -a ${homedir}/${bamclst} -o ${outdir}/${out}_13_pca.pdf
Rscript --vanilla --slave ${homedir}/pca_makeplot.R -i ./angsd_output/${out}.covar -c 2-3 -a ${homedir}/${bamclst} -o ${outdir}/${out}_23_pca.pdf

echo "Ending Job: "
date


```

**Rscript** to Plot PCA

```{r pca_plot_R, eval=F}

#!/usr/bin/Rscript
# Usage: Rscript -i infile.covar -c component1-component2 -a annotation.file -o outfile.eps

#.libPaths("~/R/x86_64-pc-linux-gnu-library/3.2")

library(optparse)
library(ggplot2)
library(stringi)
library(viridis)

#if(!require(viridis)) { 
#  install.packages("viridis"); require(viridis)} #load / install+load installr

option_list <- list(make_option(c('-i','--in_file'), action='store', type='character', default=NULL, help='Input file (output from ngsCovar)'),
                    make_option(c('-c','--comp'), action='store', type='character', default=1-2, help='Components to plot'),
                    make_option(c('-a','--annot_file'), action='store', type='character', default=NULL, help='Annotation file with individual classification (2 column TSV with ID and ANNOTATION)'),
                    make_option(c('-o','--out_file'), action='store', type='character', default=NULL, help='Output file')
                    )
opt <- parse_args(OptionParser(option_list = option_list))

# Annotation file is in plink cluster format

#################################################################################

# Read input file
covar <- read.table(opt$in_file, stringsAsFact=F);

# Read annot file
annot <- read.table(opt$annot_file, sep=" ", header=T); # note that plink cluster files are usually tab-separated instead

# Parse components to analyze
comp <- as.numeric(strsplit(opt$comp, "-", fixed=TRUE)[[1]])

# Eigenvalues
eig <- eigen(covar, symm=TRUE);
eig$val <- eig$val/sum(eig$val);
cat(signif(eig$val, digits=3)*100,"\n");


# Write eigenvalues
#write.table(eig, file = "eigen_scores.txt", quote = FALSE)

# Plot
PC <- as.data.frame(eig$vectors)
colnames(PC) <- gsub("V", "PC", colnames(PC))
PC$River <- factor(strtrim(annot$FID,5)) # adjust/add a column of interest here from metadata
PC$Pop <- factor(annot$CLUSTER)
PC$ID <- factor(annot$IID)
PC$IDname <- factor(stringr::str_extract(annot$FID, pattern = "(?<=_)([A-Z]{2})")) # change to match your naming convention

# Write PC components

#write.table(PC, file = "PC_scores.txt", quote = FALSE)

title <- paste("PC",comp[1]," (",signif(eig$val[comp[1]], digits=3)*100,"%)"," / PC",comp[2]," (",signif(eig$val[comp[2]], digits=3)*100,"%)",sep="",collapse="")

x_axis = paste("PC",comp[1],sep="")
y_axis = paste("PC",comp[2],sep="")

ggplot() + geom_point(data=PC, aes_string(x=x_axis, y=y_axis, color="IDname", shape="River"), size=5) + 
        geom_text(data=PC, aes_string(x=x_axis, y=y_axis, label="IDname", vjust= -0.784), check_overlap=TRUE, cex=2.5, hjust=-0.3) + 
        ggtitle(title) + theme_bw() + scale_color_viridis(discrete=TRUE)
ggsave(opt$out_file)
unlink("Rplots.pdf", force=TRUE)

```

## Dynamic PCA 

For data exploration, you can use the following script to make an interactive or dynamic plot, which is useful for investigating your data. You'll need to grab the `.covar` file, and have the `.clst` file from your original call.

```{r pca_plotly, echo=T, eval=T}

# LIBRARIES ---------------------------------------------------------------

suppressPackageStartupMessages({
  library(tidyverse); # for ggplot2 and dplyr
  library(viridis); # for color palettes
  library(plotly) # interactive plotting
})

# READ INPUT FILES --------------------------------------------------------

covarfile<-"rabo_mfa_100k"
clstfile<-"rabo_mfa_100k"

# read covar file
covar_path <- paste0("./data/",covarfile, ".covar")
covar <- read.table(covar_path, stringsAsFactors = F);

# Read annot file
annot <- read.table(paste0("data/bamlist_mrg_",clstfile, "_clst"),stringsAsFactors = F, header = TRUE); 

# CREATE EIGEN MATRIX -----------------------------------------------------

# Eigenvalues
eig <- eigen(covar, symm=TRUE);
eig$val <- eig$val/sum(eig$val);

# SET UP PCA -------------------------------------------------------------

# get PC colnames and set label/col/pops
PC <- as.data.frame(eig$vectors)
colnames(PC) <- gsub("V", "PC", colnames(PC))
PC$Region <- factor(annot$FID)
PC$Pop <- factor(annot$CLUSTER)
PC$ID <- factor(annot$IID)

# SET UP PLOTS ------------------------------------------------------------

# set the xy vars:
x1 = 'PC1'
y1 = 'PC2'

# Create the titles for each PCA
title12 <- paste("PC1 (",signif(eig$val[1], digits=3)*100,"%)"," / PC2 (",signif(eig$val[2], digits=3)*100,"%)", sep="",collapse="")
title13 <- paste("PC1 (",signif(eig$val[1], digits=3)*100,"%)"," / PC3 (",signif(eig$val[3], digits=3)*100,"%)", sep="",collapse="")
title23 <- paste("PC2 (",signif(eig$val[2], digits=3)*100,"%)"," / PC3 (",signif(eig$val[3], digits=3)*100,"%)",sep="",collapse="")

# TEST PLOTLY PLOTS ------------------------------------------------------

# 1 v 2
(p1v2 <- ggplotly(p = ggplot(data=PC, aes(x=PC1, y=PC2, color=Pop, 
                                          label=paste(Region,"-",ID))) +
                    geom_point(size=4, alpha=0.8) + 
                    labs(title=paste0(covarfile, ": ",title12)) +
                    theme_bw() +
                    scale_color_viridis(discrete = TRUE, option = "D")))

```


<!-- add additional code/examples here, Ryan will add additional PCA code and custom R script here-->